---
title: "Regression Project - MPG Analysis for Motor Trend"
author: "Conrad Wong"
date: "November 7, 2015"
output: html_document
---

# Executive Summary

###This analysis will use the mtcars dataset to analyze if there is a difference in MPG between manual and automatic cars.  These are the steps to be followed:

### - Data transformation and exploratory data analysis to get a sense of the shape and relationship between the different variables
### - Find the regression model that maximizes percentage of explained variance in MPG
### - Drive conclusions for two questions:
- Is an automatic or manual transmission better for MPG?
- Quantify the MPG difference between automatic and manual transmissions


# Data transformation and exploratory data analysis

```{r}

library(ggplot2)
library(GGally)
library(car)
library(knitr)

data <- mtcars

data$trans <- as.factor(ifelse(mtcars$am==0, "Auto", "Manual"))
data <- data[, -9]
data$vs <- as.factor(ifelse(data$vs==0, "V", "Straight"))
data$cyl <- as.factor(data$cyl)
data$carb <- as.factor(data$carb)
data$gear <- as.factor(data$gear)

str(data)

```


### Key Findings: The pairs plot (Figure 1 in the Appendix) suggests there might be some colinearity between the variables in the mtcars dataset, as there are some pairs that are correlated, for example: 
- disp and weight with corr= 0.89
- disp and hp with corr = 0.79

### In addition, the boxplot (Figure 2 in the Appendix) suggests manual cars have higher mpg than automatic cars.

# Model selection

## Model 1: Fit of mpg by transmission

```{r, echo=TRUE}

fit.by.trans <- lm(mpg~trans, data=data)
summary(fit.by.trans)

```

### Key Findings: Fitting by type of transmission, results in a statistical significant coefficient (p-value < 0.05), but the adjusted r2 is only 0.34.  Need to look for other variables that help explain a bigger percentage of the variance in mpg.  Next step is to fit a model with all variables

## Model 2: Fit of mpg by all variables in the mtcars dataset

```{r, echo=TRUE}
fit.by.all <- lm (mpg~. , data=data)
summary(fit.by.all)

vif(fit.by.all)

```

### Key Findings: Fitting by all the variables in the mtcars dataset increases the adjusted r2 to 0.78, but non of the coefficients is statistical significant (all p-values are greater than 0.05).  Looking at the VIF values suggests that there is colinearity between the variables.  Next step is to remove the ones with a high GVIF^(1/(2*Df)) value.

## Model 3: Fit of mpg by all variables where GVIF^(1/(2*Df)) <= 3.15

```{r, echo=TRUE}

#            GVIF Df GVIF^(1/(2*Df))
#drat    6.809663  1        2.609533
#vs      8.088166  1        2.843970
#gear   50.852311  2        2.670408
#carb  503.211851  5        1.862838
#trans   9.930495  1        3.151269

fit.by.some <- lm(mpg~ drat + vs + gear + carb + trans, data=data )
summary(fit.by.some)

vif(fit.by.some)[,3]^2

```

### Key Findings: The adjusted r2 decreases to 0.77, and we still have the problem of non statistical significant coefficients.  Next step is to fit the model by a stepwise algorithm.

## Model 4: Fit of mpg following the stepwise algorithm

```{r, echo=TRUE, warning=FALSE, message=FALSE}

stepModel <- step(fit.by.all, k=log(dim(data)[1]), trace=FALSE)
summary(stepModel)
```

### Key Findings: This approach results in the hightest adjusted r2: 0.83, and it finds three coefficients that are statistical significant (wt, qsec and transmission).  Next step is to run some diagnostics for this model

## Diagnosis and residuals of stepwise model

### Please refer to Figure 3 in the Appendix

### Key Findings: There doesn't seem to be a pattern in the residuals, so the model is a good fit.  It would be good to work with a subject matter expert to look at the lower and upper residuals in the QQ plot, and understand why they have a slight deviation from the normal distribution.


```{r, echo=TRUE}
#dfb <- round(dfbetas(stepModel)[, 4], 3)

#Edfb <- as.data.frame(sort(dfb, decreasing=TRUE))
#colnames(dfb) <- c("Transmission")

#kable(dfb)

```



# Conclusion

## Is an automatic or manual transmission better for MPG?

### - This analysis confirms that manual cars deliver more MPG than automatic cars 

## Quantify the MPG difference between automatic and manual transmissions


```{r, echo=TRUE}
coef <- summary(stepModel)$coefficients
coef[4,1] + c(-1, 1) * qt(.975, df = stepModel$df) * coef[4, 2]
```

### - With 95% confidence, we estimate that manual cars deliver between 0.05 and 5.82 more MPG than automatic cars.

# Appendix

## Figure 1: GGPairs

``` {r, echo=TRUE, warning=FALSE, message=FALSE, fig.width=12, fig.height=8}
ggpairs(data[, 2:11], lower=list(continuous="smooth"))
```

## Figure 2: Plot of MPG by Transmission

``` {r, echo=TRUE, warning=FALSE, message=FALSE, fig.width=12, fig.height=8}
plot (mpg~trans, data=data)
```

## Figure 3: Residuals

```{r, echo=FALSE}
par(mfrow=c(2,2))
plot(stepModel)

```